---
title: "Alpha diversity"
---

```{r}
#libraries
library(phyloseq)
library(vegan)
library(dplyr)
library(ggplot2)
library(svglite)
```

```{r}
# seed and pathways
set.seed(0201)
sessionInfo()
filepath = "/R_Analysis_MMS-FTS"
filepath_results =paste0( "/R_Analysis_MMS-FTS/results/", Sys.Date(), "_")
filepath_figures = paste0("/R_Analysis_MMS-FTS/figures/" , Sys.Date(), "_")
source(paste0(filepath, "/functions/plot.settings.R"))
```

```{r}
#attach data set
exp_nor = readRDS(file = paste0(filepath,"/results/2022-01-11_zOTU_dataset_scaled_rarified.rds") ) 
```

```{r}
# calculate alpha diversity of order 0-2 for the full data set
t_OTU_table = t(as.data.frame(exp_nor@otu_table))
hillnumbers = vegan::renyi(t_OTU_table, scales = c(0,1), hill = TRUE) #calculates diversity of different orders for the data set 
colnames(hillnumbers) =  c("Order0", "Order1")
hillnumbers$evenness =  hillnumbers$Order1/hillnumbers$Order0
#add the hill numbers to the environmental variables for easy plotting. 
sample_data(exp_nor) <- base::cbind(sample_data(exp_nor), hillnumbers) #add the hill numbers to the environmental variables
df_alpha = data.frame(exp_nor@sam_data)
df_alpha$samplename = rownames(df_alpha)
df_alpha = reshape2::melt(df_alpha, id.vars = colnames(df_alpha)[c(1:12,16)])
colnames(df_alpha)[14:15] = c("Order", "Diversity_number")
```

# alpha diveristy plot
```{r}
df_alpha %>% 
  dplyr::filter(Order != "evenness") %>% 
  mutate(DPH  = as.character(DPH)) %>% 
  ggplot2::ggplot(aes(x = Start_system, y = Diversity_number, group = DPH)) + 
  my.theme + 
  scale_color_manual("Rearing treatment",  values = system_overall_color) + 
  scale_fill_manual("Rearing treatment",values = system_overall_fill) + 
  scale_shape_manual("DPH",values = system_shape_day) +  
  facet_wrap(~Order, scales = "free" ) +
  geom_point(size = 3, position = position_dodge(width = 0.3), 
             aes(shape = DPH, fill = Overall_system), color = "black") + 
  geom_point(size = 2.5,  position = position_dodge(width =0.3), 
             aes(color = Overall_system, shape = DPH)) + 
  geom_point(color = "black", fill = NA, size = 3,  position = position_dodge(width = 0.3), 
             aes(shape = DPH)) + 
  guides(colour = guide_legend(override.aes = list(shape = 21)), 
         shape = guide_legend(nrow = 2)) + 
  coord_cartesian(expand = TRUE) + 
  theme(legend.position = "bottom") + 
  ylab("Diversity number") + 
  xlab("Initial treatment") + 
  theme(text = element_text(size = 14), 
        legend.position = "right")

ggsave(filename = paste0(filepath_figures, "alpha-diveristy-ord01_rarified.svg"), width = 180, height = 100, units = "mm")
```

#Difference 1 DPH between FTS and MMS 
```{r}
df_alpha_1DPH = df_alpha %>% mutate(DPH = as.character(DPH)) %>%
  group_by(Order,Start_system, Current_water, DPH) %>% 
  filter(DPH == 1) 

df_alpha_1DPH %>%  
  summarise(mean_div = round(mean(Diversity_number), 0))

order1_ratio = 39/25 #MMS/FTS  1.56
order0_ratio = 260/202 #MMS/FTS 1.287129
```

##Anova differences diveristy between water system 1 DPH.
```{r}
df_order0.1 = df_alpha %>% filter(Order == "Order0", DPH == 1)
fligner.test(data = df_order0.1, Diversity_number ~ Current_water)
shapiro.test(df_order0.1$Diversity_number)

# Compute the analysis of variance
res.aov.0.1 = aov(Diversity_number ~ Current_water, data = df_order0.1)
# Summary of the analysis
summary(res.aov.0.1)
```

```{r}
df_order1.1 = df_alpha %>% filter(Order == "Order1", DPH == 1)
fligner.test(data = df_order1.1, Diversity_number ~ Current_water)
shapiro.test(df_order1.1$Diversity_number)

# Compute the analysis of variance
res.aov.1.1 = aov(Diversity_number ~ Current_water, data = df_order1.1)
# Summary of the analysis
summary(res.aov.1.1)
```

#Carrying capacity differences 
```{r}
df_alpha %>% 
  dplyr::filter(Order != "evenness") %>% 
  mutate(DPH  = as.character(DPH)) %>% 
  group_by(Order, cc) %>% 
summarise(mean_div = mean(Diversity_number), n = n()) 

#Overall difference
290.37500/220.25000 #1.318388
50.11565/31.36190 # 1.597979

df_alpha %>% 
  dplyr::filter(Order != "evenness") %>% 
  mutate(DPH  = as.character(DPH)) %>% 
  group_by(Order, DPH, cc) %>% 
summarise(mean_div = mean(Diversity_number), n = n()) 
328.50000/231.37500	#1.419773
67.41259/ 31.25208	#2.157059	
```

#Difference 12 DPH between FTS and MMS 
```{r}
df_alpha_12DPH = df_alpha %>% mutate(DPH = as.character(DPH)) %>%
  group_by(Order, Current_water, DPH) %>% 
  filter(DPH == 12) 


df_alpha_12DPH %>%  
  summarise(mean_div = round(mean(Diversity_number), 0))

order1_ratio = 48/51 #MMS/FTS  0.9411765
order0_ratio = 291/269 #MMS/FTS 1.081784
```

##Anova differences diveristy between water system 1 DPH.
```{r}
df_order0.1 = df_alpha %>% filter(Order == "Order0", DPH == 1)
fligner.test(data = df_order0.1, Diversity_number ~ Current_water)
shapiro.test(df_order0.1$Diversity_number)

# Compute the analysis of variance
res.aov.0.1 = aov(Diversity_number ~ Current_water, data = df_order0.1)
# Summary of the analysis
summary(res.aov.0.1)
```

```{r}
df_order1.1 = df_alpha %>% filter(Order == "Order1", DPH == 1)
fligner.test(data = df_order1.1, Diversity_number ~ Current_water)
shapiro.test(df_order1.1$Diversity_number)

# Compute the analysis of variance
res.aov.1.1 = aov(Diversity_number ~ Current_water, data = df_order1.1)
# Summary of the analysis
summary(res.aov.1.1)
```

## Anova cc day 12
```{r}
df_order1.12 = df_alpha %>% filter(Order == "Order1", DPH == 12)
fligner.test(data = df_order1.12, Diversity_number ~ cc) #not sig
shapiro.test(df_order1.12$Diversity_number) #not sig

res.aov.1.12 = aov(Diversity_number ~ cc, data = df_order1.12)
# Summary of the analysis
summary(res.aov.1.12) # p 6.83e-06

df_order0.12 = df_alpha %>% filter(Order == "Order0", DPH == 12)
fligner.test(data = df_order0.12, Diversity_number ~ cc) #not sig
shapiro.test(df_order0.12$Diversity_number) #not sig

res.aov.0.12 = aov(Diversity_number ~ cc, data = df_order0.12)
# Summary of the analysis
summary(res.aov.0.12) # p 7.25e-05

```

average percent increase on start sytem level
```{r}
df_alpha_increase = df_alpha %>% filter(Order != "evenness")

meandiv_combined = df_alpha_increase %>% group_by(Order, Start_system, DPH) %>% summarise(mean_div = mean(Diversity_number)) %>% 
  mutate(DPH = paste0("Day_",DPH))

reshape2::dcast(meandiv_combined, Order+Start_system~DPH) %>% 
 mutate(ratio = round(Day_12/Day_1 , 2) ) %>%
  ggplot(aes(x=Start_system, y = ratio)) + 
  geom_bar(stat = "identity") + 
  facet_grid(~Order) +
  my.theme + 
  geom_label(aes(label = paste0(ratio)), position=position_dodge(width=0.9), vjust=0.5) + 
  geom_hline(yintercept = 0, color = "black") + 
  coord_cartesian(expand = TRUE) +
  ggtitle("Average percent increase in diversity given initial rearing regime")
```

#x fold change 1 vs 12 dph overall system average
```{r}
df_alpha_increase = df_alpha %>% filter(Order != "evenness")

meandiv_combined = df_alpha_increase %>% group_by(Order, Overall_system, Start_system, DPH) %>% 
  summarise(mean_div = mean(Diversity_number), n = n()) %>% 
  mutate(DPH = paste0("Day_",DPH))

reshape2::dcast(data = meandiv_combined, formula = Order+Overall_system+Start_system~DPH, value.var = "mean_div") %>% 
  tidyr::fill(Day_1) %>% #fill in start values for the switched by coping row above
  mutate(ratio = round(Day_12/Day_1 , 2) ) %>% #look at ratio from day 12 to 1 
  ggplot(aes(x=Overall_system, y = ratio)) + 
  geom_hline(yintercept = 1, color = "red") + 
  geom_bar(stat = "identity") + 
  ggh4x::facet_nested(Order~Start_system, scales = "free_x") +
  my.theme + 
  geom_label(aes(label = paste0(ratio)), position=position_dodge(width=0.9), vjust=0.5) + 
  geom_hline(yintercept = 0, color = "black") + 
  coord_cartesian(expand = TRUE) + 
  ylab("Ratio 12 DPH / 1 DPH") + 
  xlab("Rearing treatment")

ggsave(filename = paste0(filepath_figures, "Average percent increase in diversity given initial rearing regime.svg"), width = 180, height = 100, units = "mm")
```

#x fold change 1 vs 12 dph tank level 
```{r}
tank_div = df_alpha_increase %>% 
  group_by(Order, Start_system,Current_system, DPH, Tank) %>% 
  mutate(DPH = paste0("Day_",DPH))

tank_ratio = reshape2::dcast(data = tank_div,formula =  Order+Tank~DPH, value.var = "Diversity_number") %>% 
  mutate(ratio = round(Day_12/Day_1 ,2)) 

tank_meta  = df_alpha_increase %>% filter(DPH == 12) %>% 
  group_by(Tank, Overall_system, Start_system, Current_system ) %>% 
  summarise(n = n()) 

tank = merge(tank_meta, tank_ratio) %>% group_by(Order, Overall_system, Start_system, Current_system)  

tank %>%   
  ggplot(aes(y=Start_system, x = ratio, fill = Overall_system, group = Tank)) + 
    geom_vline(xintercept = 1, color = "red") + 
  geom_bar(stat = "identity", position = position_dodge(), color = "black", orientation = "y") + 
  facet_grid(~Order) +
  my.theme + 
  geom_text(aes(label = paste0(ratio)), position=position_dodge(1), 
            vjust=0.5, hjust = -0.1,angle = 0,  size = 4) + 
  coord_cartesian(expand = TRUE,xlim = c(0,10)) + 
  scale_fill_manual(values = system_overall_fill) 
```
