---
title: "DeSeq2"
---
```{r}
#libraries
library(DESeq2) # BiocManager::install("DESeq2")
library(phyloseq)
library(dplyr)
library(ggplot2)
library(tidyr)
#to install microbiome need to install these first
  #install.packages("adespatial")
  #BiocManager::install("WGCNA")
  #library(WGCNA)
  #library(devtools)
  #devtools::install_github("umerijaz/microbiomeSeq")
library(microbiomeSeq)
```

```{r}
# seed and pathways
set.seed(0201)
sessionInfo()
filepath = "~/Library/CloudStorage/OneDrive-NTNU/Doktorgrad/Ferdige prosjekter/Project 3 MMS_FTS_r_K_experiment/MMS-FTS Overview results and manuscript plan/R_Analysis_MMS-FTS"
filepath_results =paste0( filepath, "/results/", Sys.Date(), "_")
filepath_figures = paste0(filepath, "/figures/" , Sys.Date(), "_")

source(paste0(filepath, "/revised_figures_scripts/plot.settings.R"))
source(paste0(filepath, "/revised_figures_scripts/min.mean.sd.max.R"))
```

```{r}
#attach data set
exp = readRDS(file = paste0(filepath,"/data/zOTUs/2022-01-11_zOTU_raw_dataset.rds") )
#add survival data to metadata
survival = readRDS(paste0(filepath, "/results/2022-01-14_survival.RDS"))
survival = survival %>% select(tank_letter,survival ) 
colnames(survival)[1] = "Tank"
survival_df = phyloseq::sample_data(merge(data.frame(sample_data(exp) ), survival))
rownames(survival_df) = survival_df$Sample
sample_data(exp) = survival_df
```

```{r}
physeq = exp
tax_table(physeq) = cbind(tax_table(physeq), rownames(tax_table(physeq)))
colnames(tax_table(physeq)) = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "OTUID")
#want to compare low and high survival. Setting treshold at 10% survival
sample_data(physeq)$sur_cat = case_when(sample_data(physeq)$survival > 10 ~ "moderate", 
                                         sample_data(physeq)$survival < 10 ~ "low") 
physeq = subset_samples(physeq = physeq, DPH == 12)
```

Following guide given in Statistical Analysis of Microbiome Data with R chapter 11.5 The DESeq and DESeq2 Packages
https://doi.org/10.1007/978-981-13-1534-3

Step 1: Create the Count Table
```{r}
# convert OTU table to matrix
otu_tab = (otu_table(physeq))
#DESeq2 need taxa(genes=rows) by samples(=columns)
countData<-(as(otu_tab, "matrix"))
#head(countData)
```

Step 2: Create the Sample Metadata Table
We want moderate to be "control" level.
```{r}
#sample metadata
meta = data.frame(sample_data(physeq))
meta$sur_cat = factor(meta$sur_cat, levels = c("moderate", "low"))
group = meta$sur_cat
```

Step 3: Build the DESeq2 Object
```{r}
# build the deseq object 
metaData = data.frame(row.names=colnames(countData),group=group)
dds = DESeqDataSetFromMatrix(countData = countData, colData = metaData, design = ~group)
```

Step 4: Filter the Data
```{r}
dds <- dds[rowSums(counts(dds)) > 0,] #double check all 0 count taxa are gone
#normalise data with median ration method.
```

Step 5: Normalize the Count Data
```{r}
dds <- estimateSizeFactors(dds)
#sizeFactors(dds)
```

Step 6: Estimate the Dispersion
```{r}
#estimate the dispersion 
dds<- estimateDispersions(dds)
```

Step 7: Test the Differential Abundance
```{r}
deseq_restult = DESeq2::DESeq(object = dds, test = "Wald", fitType = "parametric" )
#saveRDS(object = deseq_restult, file = paste0(filepath_results, "DeSeq_object_survival.RDS"))
```

```{r}
res <- results(deseq_restult)
```

```{r}
# how many OTUs were sig differentailly expressed (using FDR ajusted p-value)
table(res[,"padj"] < 0.1) # the FDR corrected p-value have 192 below 0.1
table(res[,"padj"] < 0.01) # 75  below 0.01
table(res[,"padj"] < 0.05) # 137 below 0.05
```

Investigate the fold changes distribution along with the p-values through a volcanoplot. 
```{r}
df = as.data.frame(res) 
df$taxa = rownames(df) 
df1 = na.omit(df)
df1 %>% ggplot2::ggplot(aes(x= log2FoldChange, y = -log10(padj))) + 
  geom_point()
```

Create a dataframe with the significant OTUs
```{r}
alpha = 0.05
sigtab = res[which(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(physeq)[rownames(sigtab), ], "matrix"))
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))

# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))
sigtab2 = sigtab %>% mutate(classified = substr(Order, 1,1)) %>% 
  mutate(Ordername =case_when(classified == "o" ~ substring(Order, 3), 
                              classified != "o" ~ "x_unclassified at order level")) %>% 
  mutate(classified2 = substr(Genus, 1,1)) %>% 
  mutate(taxonname = case_when(classified2 == "g" ~ substring(Genus,1), 
                              classified2 != "g" ~  substring(Genus, 6) )) %>%
  mutate(lowest_class = substr(taxonname, 1,1)) %>%
  mutate(group = 
           case_when(classified2 == "g" ~  paste0(substring(taxonname,3)), 
                     classified2 != "g" ~  paste0("uncl. (", substring(taxonname,1,1), " ", substring(taxonname,3), ")"))) %>%
  mutate(classified_c = substr(Class, 1,1)) %>% 
  mutate(classname =case_when(classified_c == "c" ~ substring(Class, 3), 
                              classified_c != "c" ~ "x_unclassified at class level")) %>%
  mutate(classified_f = substr(Family, 1,1)) %>% 
  mutate(classname_f =case_when(classified_f == "f" ~ substring(Family, 3), 
                              classified_f != "f" ~ "x_unclassified at family level"))
saveRDS(object = sigtab2, file = paste0(filepath_results, "DeSeq_alpha0-05_survival_taxa_12DPH.RDS"))
```


```{r}
sigtab2 = readRDS(file = paste0(filepath, "/results/2022-01-18_DeSeq_alpha0-05_survival_taxa_12DPH.RDS"))

sigtab2 %>% 
  ggplot2::ggplot(aes(x=log2FoldChange, y = Genus, fill = Order)) + 
  geom_point(shape = 21, size = 2) + 
  my.theme + 
  geom_vline(xintercept = 0) + 
  ylab("") + 
  scale_fill_manual(values = col[1:21]) + 
  theme(text = element_text(size = 12)) + 
    guides(fill = guide_legend(title.position = "top", override.aes = list(shape = 21, size = 3)) )

ggsave(filename = paste0(filepath_figures, "deseq2_staende_order.svg"), 
       width = 180*1.2, height = 150, units = "mm" ) 
```

# Taxa from DeSeq2 potting
```{r}
deseq2_taxa = readRDS(file = paste0(filepath, "/results/2022-01-18_DeSeq_alpha0-05_survival_taxa_12DPH.RDS"))
my.physeq =readRDS(file = paste0(filepath,"/results/2022-01-11_zOTU_dataset_scaled_rarified.rds") )
source(paste0(filepath, "/revised_figures_scripts/plot.settings.R"))
source(paste0(filepath, "/revised_figures_scripts/min.mean.sd.max.R"))
source(paste0(filepath, "/revised_figures_scripts/summarize_taxa_functions.R"))
tax_table(my.physeq) = cbind(tax_table(my.physeq), rownames(tax_table(my.physeq)))
colnames(tax_table(my.physeq)) = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "OTUID")
deseq2_taxa_linked = deseq2_taxa %>% mutate (linked_to = case_when(log2FoldChange > 0 ~ "low survival", 
                                              log2FoldChange < 0 ~ "moderate survival")) %>% select(OTUID,linked_to )
```

```{r}
new_meta = readRDS(file = paste0(filepath, "/results/2021-06-30_new_metadata.RDS") ) 
sample_data(my.physeq) = new_meta
sample_data(my.physeq)$sur_cat = case_when(sample_data(my.physeq)$survival > 10 ~ "moderate", 
                                         sample_data(my.physeq)$survival < 10 ~ "low") 
# for plotting purposes we want to work with compositional data (scale in percentage)
my.physeq = microbiome::transform(x = my.physeq, transform = "compositional")

tax_rank = "OTUID"
min_percentage = 0
minTotRelAbun = min_percentage/100
nsamples_above = 0 

df_keep_taxa = data.frame(summarize_taxa(physeq = my.physeq, Rank = tax_rank, GroupBy = "Sample"))
df_keep_taxa = df_keep_taxa %>% group_by(df_keep_taxa[1]) %>% 
  summarise(in_n_samples = n(), 
            n_samples_above_min_ab = sum(totalA > minTotRelAbun ) ) %>%
  filter(n_samples_above_min_ab >= nsamples_above )

taxa_summary = data.frame(summarize_taxa(my.physeq, tax_rank, GroupBy = c("Sample")))
taxa_summary_df = merge(taxa_summary, data.frame(sample_data(my.physeq)), by = "Sample")
taxa_summary_df_filtered = taxa_summary_df %>% filter(OTUID %in% deseq2_taxa$OTUID ) 
```
```{r}
taxa_plotting = deseq2_taxa %>% select(Phylum, Class, Order, Family, Genus, OTUID, group, classname, classname_f)
taxa_summary_df_filtered.1 = merge(taxa_summary_df_filtered,taxa_plotting)
taxa_summary_link = merge(taxa_summary_df_filtered.1 , deseq2_taxa_linked) 

taxa_summary_link.1 = taxa_summary_link %>% select(OTUID, Sample, totalA, Tank, DPH, Overall_system, Start_system, Class, Order, Family, Genus, classname_f,linked_to, sur_cat)

combinations =  taxa_summary_link.1  %>% expand(
  nesting(OTUID,Class, Order, Family, Genus, classname_f, linked_to),  
  nesting(Sample, Tank, DPH, Overall_system, Start_system,sur_cat) )  

gaps_filled = taxa_summary_link.1 %>% dplyr::right_join(combinations) %>% 
  mutate(rep = paste0(Overall_system, "  " ,Tank))
```

```{r}
taxa_summary_link_df = 
  gaps_filled %>%
  
  na.omit() %>% 
  group_by(Sample, Overall_system, linked_to, Start_system, Order, rep, DPH) %>% 
  summarise(tot_a = sum(totalA)) %>%
  mutate(link_label = paste0("Linked to ", linked_to) ) 

total_per_sample = taxa_summary_link_df %>% group_by(rep, DPH, Overall_system) %>% summarise(tot_a = sum(tot_a)) 
total_per_sample$Order = "x_Total abundance"
total_per_sample$link_label = "Total abundance"


total_per_sample_df = bind_rows(taxa_summary_link_df, total_per_sample)
```

```{r}
total_per_sample_df %>%  
  ggplot2::ggplot(aes(y=rep, x = tot_a)) +
  
  ggh4x::facet_nested(DPH  + Overall_system ~link_label ,switch = "y", 
             space = "free_y", 
             scales = "free_y") + ylab("") + 
  geom_bar(stat = "identity", aes(fill = Order), color = "black" ) + 
  scale_x_continuous(labels = scales::percent_format(accuracy = 1, suffix = "")) + 
  my.theme + 
  scale_fill_manual(values = c(col[1:16], "grey90") ) + 
  xlab("Abundance [%]") + coord_cartesian(expand = F) + 
  
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),  
        strip.text.y.left = element_text(angle = 0), 
        panel.spacing.x   = unit(0.2, "lines"),
        text = element_text(size = 8), 
  panel.spacing.y   = unit(0, "lines")) 
```


```{r}
ggsave(filename = paste0(filepath_figures, "deseq2_abundance_each_sample_composition_w-total.svg"), 
       width = 180, height = 130, units = "mm" )
```



Number of ASVs linked to low and moderate survival 
```{r}
deseq2_taxa_linked %>% group_by(linked_to) %>% summarise(n = n())
```

# g_Moritella
```{r}
mortinella = gaps_filled %>%   mutate(link_label = paste0("Linked to ", linked_to) ) %>%
    filter(Genus == "g_Moritella") %>% na.omit() %>% 
  group_by(OTUID) %>% summarise(n=n())

exp_nor = readRDS(file = paste0(filepath,"/results/2022-01-11_zOTU_dataset_scaled_rarified.rds") )

exp_g_Moritella = subset_taxa(exp_nor, Genus == "g_Moritella")

View(exp_g_Moritella@otu_table)
plot_bar(exp_g_Moritella, fill = "Genus") + facet_grid(~Overall_system+DPH, space = "free", scales = "free")
```

```{r}
gaps_filled %>%   mutate(link_label = paste0("Linked to ", linked_to) ) %>%
    filter(Genus == "g_Moritella") %>% na.omit() %>% 
  group_by(OTUID) %>% summarise(n=n())
```


